\documentclass[11pt,a4paper]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{xcolor}

%% placeholder marker for anything still to be filled in
\newcommand{\TODO}[1]{\textcolor{red}{[\textbf{TODO:} #1]}}

\title{Sicilian, English and Italian Neural Machine Translation:\\
       an Automated Data Pipeline and Reproducible Baselines}
\author{Donato Meoli}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
Sicilian is a low-resource Romance language, with a literary tradition older than Italian,
yet absent from every major machine-translation service. Building on Wdowiak's
\emph{Tradutturi Sicilianu} \citep{wdowiak2021recipe}, the first Sicilian neural machine
translation (NMT) system, we rebuild the parallel-corpus pipeline with sentence-embedding
extraction and alignment (PyMuPDF and LaBSE \citep{feng2022labse}), assemble a reproducible,
deduplicated Sicilian--English dataset with a held-out \emph{literary} test set, and port the
paper's tokenization and desinence-biasing recipe to a Perl-free stack. We evaluate both
from-scratch Sockeye baselines and the pretrained NLLB-200 model \citep{nllb2022} adapted with
LoRA \citep{hu2021lora}. On our harder held-out test, Sicilian$\rightarrow$English climbs from
a 5.27 BLEU floor to 31.43 with a bidirectionally fine-tuned NLLB-1.3B---above Wdowiak's
published baseline (29.1)---while the reverse direction nearly doubles (9.89$\rightarrow$18.73).
We further include \textbf{Italian} as a closely-related, higher-resource bridge language
(\texttt{ita\_Latn}): NLLB supports it$\leftrightarrow$scn natively, and we evaluate and
fine-tune those directions on a frozen WikiMatrix it--scn split. The full pipeline, dataset
assembly and evaluation are reproducible.
\end{abstract}

\section{Introduction}
Sicilian is a Romance language spoken daily by millions in Sicily, Calabria and Puglia, and in
diaspora communities abroad. The Sicilian School of poets at the court of Frederick~II created
the first literary standard in Italy in the 13\textsuperscript{th} century and inspired Dante;
Sicilian thus emerged as a literary language \emph{before} Italian. Despite this, it is absent
from Google Translate, Bing, Yandex and DeepL---as are most regional languages.

The bottleneck was never the model but the \emph{parallel text}: nobody had assembled it. The
material exists, in the bilingual literary journal published for over forty years by Arba
Sicula. Wdowiak's \emph{Tradutturi Sicilianu} \citep{wdowiak2021recipe} was the first Sicilian
NMT system, trained on hand-aligned Arba Sicula text with a recipe tailored to the
low-resource, morphologically rich setting. This work rebuilds that effort with maintained,
mostly CPU-only tooling and pushes quality with modern pretrained models.

\paragraph{Contributions.}
\begin{itemize}
  \item An automated extraction and alignment pipeline for the \emph{Arba Sicula} PDFs (PyMuPDF
        and LaBSE \citep{feng2022labse}), replacing the legacy pdflatex+hunalign flow, with
        thresholds tuned against Wdowiak's hand-aligned gold.
  \item A reproducible, deduplicated, provenance-tracked dataset with a held-out literary test
        set (not FLORES), and a BLEU/chrF evaluation harness.
  \item The paper's ``lever B'' recipe (Napizia tokenizer + desinence-biased subwords) ported
        to a Perl-free pipeline, and a linguistic-feature lever (lemma source factors).
  \item NLLB-200 zero-shot and LoRA fine-tuning that beats the published baseline on our harder
        test set, with a bidirectional adapter that also rescues the reverse direction.
\end{itemize}

\section{Background}
NMT models translation as a conditional language model: an encoder reads the source
$x=(x_1,\dots,x_n)$ and a decoder factorises the target left to right,
$p_\theta(y\mid x)=\prod_{t} p_\theta(y_t\mid y_{<t},x)$, trained by maximising the
log-likelihood of a parallel corpus and decoded with beam search. The dominant
parameterisation is the \textbf{Transformer} \citep{vaswani2017attention}, which replaces
recurrence with scaled dot-product attention,
$\mathrm{Attention}(Q,K,V)=\mathrm{softmax}(QK^{\top}/\sqrt{d_k})V$, run in parallel heads.

With little data the default Transformer overfits; the low-resource recipe of
\citet{sennrich2019loresnmt} prescribes a smaller, heavily regularised model with high dropout
\citep{srivastava2014dropout} and small subword vocabularies. \textbf{Subword segmentation}
via byte-pair encoding \citep{sennrich2016subword} gives an open vocabulary and shares
statistics across word forms---essential for a morphologically rich language. Further gains
come from \textbf{linguistic input features} \citep{sennrich2016linguistic},
\textbf{back-translation} of monolingual data \citep{sennrich2015backtrans}, and
\textbf{multilingual} training with a target-language token \citep{johnson2017zeroshot} that
enables bridging through related high-resource languages \citep{fan2020beyond,
arivazhagan2019massively}. Low-resource MT remains hard for exactly these reasons
\citep{koehnknowles2017}.

\section{Data Pipeline}
\subsection{Extraction from PDFs}
We extract clean text from the Arba Sicula issue PDFs with PyMuPDF (no pdflatex round-trip).
Each page is assigned a language from its share of Sicilian stop-words and diacritics; the
journal sets Sicilian and English on facing pages, giving a natural even/odd prior, so we form
facing candidate pairs. Pages are segmented into sentences with \texttt{pysbd} (per language),
filtering ``furniture'' (all-caps headers, page numbers, captions). The procedure recovers
parallel pages the manual selection had skipped.

\subsection{Sentence alignment}
We embed every sentence with LaBSE \citep{feng2022labse}, a language-agnostic dual encoder, and
build the cosine-similarity matrix $S_{ij}=\langle u_i,v_j\rangle/(\lVert u_i\rVert\lVert
v_j\rVert)$. A facing pair is confirmed only if its mean best-match similarity exceeds
$\tau_{\text{page}}$; sentences are then aligned by a monotonic dynamic program over $S$
allowing $1$--$1$, $1$--$2$ and $2$--$1$ moves with a null penalty, in the spirit of Vecalign
\citep{thompson2019vecalign} but without a bilingual dictionary or hunalign. The thresholds
$(\tau_{\text{page}},\tau_{\text{sent}})=(0.50,0.40)$ were tuned against Wdowiak's hand-aligned
gold (issues AS41--42); the pipeline recovers 14{,}245 sentence pairs from 44 issues, comparable
to his manual alignment. Crucially, the \emph{unaligned} English is retained as an in-domain
monolingual pool for back-translation.

\subsection{Additional sources}
We add the author's curated ``Good-Sicilian-in-NLLB'' set---LASER3-mined sentences
\citep{heffernan2022laser3, nllb2022} rescored and filtered for quality and Corsican
contamination---a small curated ``Good-Sicilian-from-WikiMatrix'' set, and the WikiMatrix
it--scn subset \citep{schwenk2021wikimatrix} as an auxiliary Italian bridge.

\subsection{Unified dataset}
Sources are cleaned, deduplicated across one another, and split deterministically. A 1{,}000/1{,}000
valid/test split is \emph{frozen} once, held out from Arba Sicula (a literary standard, not
FLORES), so every later result stays comparable (Table~\ref{tab:data}).

\begin{table}[t]\centering
\caption{Unified dataset.}\label{tab:data}
\begin{tabular}{lrl}
\toprule
split & pairs & source \\
\midrule
train & 27{,}386 & Arba Sicula + NLLB \\
valid & 1{,}000 & Arba Sicula \\
test  & 1{,}000 & Arba Sicula (literary) \\
it--scn (aux) & 11{,}389 & WikiMatrix \\
\bottomrule
\end{tabular}
\end{table}

\section{Standardization, Tokenization and Subwords}
Sicilian has no single official orthography, so we first \textbf{normalise} text toward
\emph{Standard Sicilian}---the standard of Cipolla, augmented by Wdowiak for MT clarity: strict
\textbf{L} on articles (\emph{lu/la/li}), \textbf{h} on \emph{aviri} (\emph{haiu, havi, hannu}),
\textbf{CI} for \c ci (\emph{\c ciuri$\to$ciuri}), and \textbf{R} retained before a clitic
(\emph{farlu}). Our \texttt{normalize\_scn.py} applies the documented spelling rules
(e.g.\ \emph{cchi\`u$\to$chi\`u}, \emph{io/iu/eu$\to$jo}, \emph{non$\to$nun}) at a ``std'' level,
and additionally uncontracts and ASCII-folds at a ``full'' level. Because NLLB's own tokenizer
was trained on un-normalised web Sicilian, whether this normalisation \emph{helps} the
pretrained model is an empirical question we ablate (raw vs std vs full).

We re-implement the Napizia tokenizer in pure Python (verified identical to the original Perl):
it ASCII-folds accents and \emph{uncontracts} Sicilian (e.g.\ \^o$\to$\emph{a lu},
d\^u$\to$\emph{di lu}), cutting sparsity. ``Lever B'' then biases the joint scn--en BPE
\citep{sennrich2016subword} toward the textbook \emph{desinences} catalogued in the resources
Wdowiak assembled to standardise Sicilian---Dieli's \emph{Sicilian Vocabulary}, Cipolla's
\emph{Mparamu lu sicilianu} \citep{cipolla2013}, Bonner's grammar \citep{bonner2001}, and the
\emph{Chi\`u d\^a Palora} dictionary---so morphological endings survive as units (Wdowiak's
example: \emph{Carinisi are dogs!} segments to \texttt{car@@ in@@ isi are dogs !}). This cut
the vocabulary from 14{,}000 to 2{,}000 units and produced his largest BLEU gains. BPE-dropout
\citep{provilkov2020bpedropout} adds subword regularisation by sampling segmentations during
training.

\section{Models}
The from-scratch baseline is a Sockeye-3 \citep{hieber2017sockeye} small Transformer (3 layers,
model size 256, 4 heads, feed-forward 1024, weight tying, dropout 0.4, joint BPE $\approx$4k;
$\sim$6.6M parameters), trained on CPU. On top of it we stack lever B (tokenization +
desinences), more parallel text, and ``lever D'' linguistic input features
\citep{sennrich2016linguistic}: a lemma source factor summed into the source embedding,
$e_j=E_{\text{word}}(w_j)+E_{\text{lemma}}(\ell_j)$, whose small ($\sim$900) vocabulary
generalises across inflected forms.

The pretrained model is NLLB-200 \citep{nllb2022}, a massively multilingual encoder--decoder
that supports Sicilian (\texttt{scn\_Latn}) and is steered by forcing the first decoder token.
We adapt the 1.3B variant with LoRA \citep{hu2021lora}: freezing $W_0$ and learning a low-rank
update $W=W_0+\frac{\alpha}{r}BA$ ($r{=}32$, $\alpha{=}64$) on the attention projections, with
the base in fp16 and the adapters in fp32. Training \emph{both} directions at once (scn$\to$en
and en$\to$scn) yields a single adapter that serves both and rescues the weak reverse direction.

\section{Evaluation}
We score with sacreBLEU \citep{post2018sacrebleu}---BLEU \citep{papineni2002bleu}
(\texttt{tok:13a}) and chrF \citep{popovic2015chrf}---on the
frozen literary test set rather than the NLLB/FLORES set: the FLORES Sicilian uses the
post-2017 orthography and a different (non-literary) domain, so it is not representative of the
Arba Sicula material. The Sockeye rows are scored in tokenized space (the same space as the
upstream system; the raw-text floor is 5.27 BLEU), while the NLLB rows are scored on raw text.

\section{Results}
On the frozen held-out test set (1{,}000 literary pairs), Sicilian$\rightarrow$English results
are reported in Table~\ref{tab:results}. The Sockeye levers stack from 5.54 to 10.85 BLEU on CPU
alone; the pretrained NLLB-1.3B wins decisively even zero-shot (29.02), and LoRA fine-tuning
lifts it to 31.43---above Wdowiak's published baseline (29.1) on our harder test.

\begin{table}[t]\centering
\caption{Our models, our test set (scn$\rightarrow$en).}\label{tab:results}
\begin{tabular}{lrr}
\toprule
model & BLEU & chrF \\
\midrule
floor (copy source) & 5.27 & 25.40 \\
Sockeye baseline & 5.54 & 28.28 \\
\;+ lever B (tok.\ + desinences) & 7.24 & 29.52 \\
\;+ more data & 9.79 & 33.82 \\
\;+ lever D (lemma source factors) & 10.85 & 35.22 \\
NLLB-200 1.3B, zero-shot (raw) & 29.02 & 55.23 \\
NLLB-200 1.3B, LoRA fine-tuned, scn$\rightarrow$en only & 31.16 & 56.79 \\
NLLB-200 1.3B, LoRA \emph{bidirectional} & \textbf{31.43} & \textbf{56.94} \\
\bottomrule
\end{tabular}
\end{table}

In the reverse direction (en$\rightarrow$scn), the same bidirectional adapter lifts NLLB-1.3B
from 9.89 (zero-shot) to 18.73 BLEU / 49.96 chrF, nearly doubling the weak direction at no cost
to scn$\rightarrow$en and giving a usable two-way model. This is still below the published
en$\rightarrow$scn baseline (25.1); back-translation is the next lever.

\paragraph{Italian directions (it$\leftrightarrow$scn).}
Italian is a closely related, much higher-resource Romance language, and NLLB supports
it$\leftrightarrow$scn natively. We hold out a frozen 1{,}000-pair it--scn test split from the
WikiMatrix data, evaluate NLLB-1.3B zero-shot, then fine-tune \emph{all four} directions
jointly (scn$\leftrightarrow$en and it$\leftrightarrow$scn) into a single multilingual adapter,
a bridge in the sense of \citet{johnson2017zeroshot} and \citet{fan2020beyond}
(Table~\ref{tab:italian}). Italian is by far the easiest pair---scn$\rightarrow$it reaches
\textbf{43.5} BLEU and it$\rightarrow$scn climbs $17.6\rightarrow\textbf{27.0}$---as expected
for a close, high-resource neighbour on in-domain WikiMatrix text. The multilingual adapter
costs a little on scn$\leftrightarrow$en ($30.8/17.5$ vs the dedicated $31.4/18.7$), so one can
ship either the trilingual all-rounder or the specialised pair.

\begin{table}[t]\centering
\caption{Trilingual model: all directions, zero-shot vs.\ the multilingual fine-tune
(BLEU / chrF, frozen test sets).}\label{tab:italian}
\begin{tabular}{lcc}
\toprule
direction & zero-shot & multilingual FT \\
\midrule
scn$\rightarrow$en & 29.02 / 55.23 & 30.75 / 56.57 \\
en$\rightarrow$scn & 9.89 / 41.12 & 17.49 / 48.78 \\
it$\rightarrow$scn & 17.61 / 44.77 & \textbf{26.97 / 51.69} \\
scn$\rightarrow$it & 42.20 / 59.04 & \textbf{43.47 / 59.79} \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Relation to the state of the art.}
We do \emph{not} beat the state of the art. Wdowiak's \textbf{reverse-training} system reports,
on his own in-domain hand-selected test set, En$\rightarrow$Sc/Sc$\rightarrow$En BLEU of
45.1/48.6 and It$\rightarrow$Sc/Sc$\rightarrow$It of 61.4/62.9---well ahead of our 18.7/31.4 and
27.0/43.5 on every direction (his earlier marks were 25.1/29.1 baseline, 35.0/36.8 augmented).
The numbers are \emph{not} a head-to-head (different test sets), but the gap is too large to be
explained by that alone: on Sc$\rightarrow$en we only clear his \emph{baseline} (31.4 vs 29.1),
on a harder test, and trail elsewhere. Crucially, his reverse-training builds a \emph{custom
pretrained model} from tens of millions of pairs (forward-translation, then back-translation,
in three stages) before fine-tuning on \emph{Arba Sicula}---a pipeline infeasible to reproduce
on a single GPU. NLLB-200, however, already provides that massive multilingual pretraining: our
fine-tune is the analogue of his stage-3. Adding back-translation and multilingual training (in
progress) closes the remaining \emph{method} gap; the likely \emph{residual} is then
\textbf{data}---his hand-curated, standardised corpus versus our automatic extraction---which
argues for combining the two rather than competing.

\section{Conclusion and Future Work}
We presented a reproducible, Perl-free path from Arba Sicula PDFs to a Sicilian--English NMT
system: an automated extraction-and-alignment pipeline, the paper's tokenization and desinence
recipe with an added linguistic-feature lever, and NLLB-200 adapted with LoRA. On our harder
literary test set the best model reaches 31.43 BLEU (scn$\rightarrow$en), above the published
baseline but well below Wdowiak's reverse-training state of the art. Ongoing work applies the
feasible parts of his recipe on top of NLLB---\textbf{back-translation} from the in-domain
monolingual English harvested from unaligned text, and \textbf{multilingual} Italian bridging
through the it--scn data \citep{fan2020beyond}---to isolate how much of the gap is
\emph{method} versus \emph{data}. If, with the method matched, our automatically-extracted
corpus still trails his hand-curated one, the difference is the data: a concrete case for
combining his corpus with our pipeline. We also release \textbf{serving} (a local web app and a
Telegram bot, a FastAPI service over the NLLB adapter). All code, the data pipeline and the
evaluation harness are released for reproducibility.

\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
